---
title: "Pathways Heart Study Aim 2 IPW Analysis"
author: "Hanjie Shen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The original Pathways cohort includes 4505 participants. For this project, we restricted women to be enrolled for 12 months prior to their diagnosis date. There is no such restriction for the participants in the original Pathways cohort, therefore, when we applied the enrollment restriction, we ended up selecting 4181 women from the original Pathways cohort.

```{r, message=FALSE, warning=FALSE}
require(tidyverse)
require(haven)
require(ggplot2)
require(dplyr)
require(plyr)
require(naniar)
require(reshape2)
require(survival)
require(survminer)
require(cmprsk)
require(pec)
require(LtAtStructuR)
require(stremr)
require(magrittr)
require(grDevices)
require(data.table)
require(knitr)

# Load data
setwd("~/Desktop/Pathways Heart Study/Data/full data")
# setwd("C:/Users/shenh/Desktop/Pathways Heart Study/Data/full data")

# cases data
cases = as.data.frame(read_sas('cases.sas7bdat'))
cases_rad = as.data.frame(read_sas('aim1_cases_rad_tx_02mar2020.sas7bdat'))
cases$rad_tx_yn = cases_rad$rad_tx_yn[match(cases_rad$CVD_studyid,cases$CVD_studyid)]
cases$laterality_tx = cases_rad$LATERALITY[match(cases_rad$CVD_studyid,cases$CVD_studyid)]
cases$rad_tx_start_dt = cases_rad$rad_tx_start_dt[match(cases_rad$CVD_studyid,cases$CVD_studyid)]
cases = cases[,!names(cases) %in% c("rad_yn","rad_date","laterality")]

# risk factor data 
bmi = as.data.frame(read_sas('baseline_bmi.sas7bdat'))
bp = as.data.frame(read_sas('baseline_bp.sas7bdat'))
lipid = as.data.frame(read_sas('dyslipidemia.sas7bdat'))
diab = as.data.frame(read_sas('diabetes.sas7bdat'))
smok = as.data.frame(read_sas('smoking.sas7bdat'))
hyper = as.data.frame(read_sas('phase_hypertension.sas7bdat'))

# income and education data
educ_income = as.data.frame(read_sas('pw_educ_income.sas7bdat'))

# menopause and census data 
menop = as.data.frame(read_sas('menopause.sas7bdat'))
census = as.data.frame(read_sas('census.sas7bdat'))

# cardiotox
pw_cardiotox = as.data.frame(read_sas('pw_cardiotox.sas7bdat'))

# anthra
pw_anthra = as.data.frame(read_sas('pw_anthra_bdcomp_14jan21.sas7bdat'))

# censoring events 
censor = as.data.frame(read_sas('censoring.sas7bdat'))
censor_phs = as.data.frame(read_sas('censoring_pathways.sas7bdat'))


# chemotherapy data
chemo = as.data.frame(read_sas('aim2_chemo_data_08apr2020.sas7bdat'))
chemo_drug = as.data.frame(read_sas('aim2_chemo_drug_ind_add_11dec20.sas7bdat'))


# hormonal data
horm = as.data.frame(read_sas('aim2_horm_rx_08may2020.sas7bdat'))
names(horm) = tolower(names(horm))

# EF dataset
ef = as.data.frame(read_sas('aim2_ef.sas7bdat'))

# Charlson dataset
charlson = as.data.frame(read_sas('kwan_num0090_01jun20.sas7bdat'))

# CVD outcome data 
cvd = as.data.frame(read_sas('cvd_events.sas7bdat'))

# clean CVD data
names(cvd) = tolower(names(cvd))
# recode cvd group
cvd$cvd_group[cvd$cvd_condition=='Heart Failure'] = 'Heart Failure'
cvd$cvd_group[cvd$cvd_condition=='Cardiomyopathy'] = 'Cardiomyopathy'
cvd$cvd_group[cvd$cvd_condition=='Transient Ischemic Attack (TIA)'] = 'TIA'
cvd$cvd_group[cvd$cvd_condition %in% c('Acute Ischemic stroke',
                                       'Acute myocardial infarction',
                                       'Intracerebral hemorrhage',
                                       'Other cerebrovascular disease',
                                       'Retinal vascular occlusion',
                                       'Subarachnoid hemorrhage')] = 'Stroke'

#### define true incidence and prevalence
cvd2 = suppressWarnings(reshape2::dcast(data=cvd, cvd_studyid ~ cvd_group,
             value.var = 'daysto_cvd_event', min, na.rm=T))

cvd2[,-1] = lapply(cvd2[,-1], function(x) {
  x[x==Inf] = NA
  x
})

names(cvd2) = gsub('/| ','_',names(cvd2))
cvd_grp = paste0(names(cvd2)[-1],'_grp')
names(cvd2)[-1] = paste0(cvd_grp,'_dt')

cvd2[,cvd_grp] = lapply(cvd2[,-1], function(x) {
  y = ifelse(!is.na(x),1,0)
  y[y==1 & x>0] = 2
  y = factor(y, levels=0:2, labels = c('No','Prevalent','Incident'))
  y
})

cvd2list = lapply(cvd_grp, function(x){
  dates = eval(parse(text=paste0("dcast(data=cvd2, cvd_studyid~",x,", value.var = '",x,"_dt')")))
  prev = ifelse(cvd2[,x]=='Prevalent',1,0)
  inc = ifelse(cvd2[,x]=='Incident',1,0)
  sum = data.frame(cbind(prev,inc,dates))
  sum = sum[,c('cvd_studyid','prev','inc','Incident')]
  names(sum) = c('cvd_studyid',paste0(x,c('_prev','_inc','_incdt')))
  sum
})


cvd2list[[12]] = cvd2
cvd2_final = Reduce(function(...) merge(...,by='cvd_studyid', all.x=T), cvd2list)


#### define true incidence and prevalence
cvd3 = suppressWarnings(reshape2::dcast(data=cvd, cvd_studyid~cvd_condition, 
             value.var = 'daysto_cvd_event', min,na.rm=T))


names(cvd3) = gsub('/| ','_',names(cvd3))
names(cvd3) = gsub('&_|\\(|\\)','',names(cvd3))
names(cvd3)[17] = "Percutaneous_coronary_intervention_PCI"

cvd3[,-1] = lapply(cvd3[,-1], function(x) {
  x[x==Inf] = NA
  x
})

cvd_cond = names(cvd3)[-1]
names(cvd3)[-1] = paste0(names(cvd3)[-1],'_dt')
cvd3[,cvd_cond] = lapply(cvd3[,-1],  function(x) {
  y = ifelse(!is.na(x),1,0)
  y[y==1 & x>0] = 2
  y = factor(y, levels=0:2, labels = c('No','Prevalent','Incident'))
  y
})

cvd3list = lapply(cvd_cond, function(x){
  dates = eval(parse(text=paste0("dcast(data=cvd3, cvd_studyid~",x,", value.var = '",x,"_dt')")))
  prev = ifelse(cvd3[,x]=='Prevalent',1,0)
  inc = ifelse(cvd3[,x]=='Incident',1,0)
  sum = data.frame(cbind(prev,inc,dates))
  sum = sum[,c('cvd_studyid','prev','inc','Incident')]
  names(sum) = c('cvd_studyid',paste0(x,c('_prev','_inc','_incdt')))
  sum
})

cvd3list[[23]] = cvd3
cvd3_final = Reduce(function(...) merge(...,by='cvd_studyid', all.x=T), cvd3list)

names(cvd2_final)
names(cvd3_final)


## list all datasets
datalist = list(chemo[,c(-5,-6,-12,-13,-14)],
                 cases[,-1],
                 bmi[,c("CVD_studyid","BMI","BMI_Category_II","daysto_bmi")],
                 bp[,c("CVD_studyid","systolic","diastolic","daysto_bp","hypertension")],
                 diab[,c(6,7,1)],lipid[,c(4,5,2)], hyper[,c(6,2,7)],
                 smok[,c(5,2)],menop[,c(1,3)],
                 census[,c(4,5,6,7,8,9,10,11,12,13,14,15,16,2)],
                 cvd2_final,
                 cvd3_final,
                 censor[,c(1,9,10,11,12,13)],
                 censor_phs,
                 chemo_drug,
                 pw_anthra[,-4])

## make all var names lower case
datalist = lapply(datalist,function(x){
  names(x) = tolower(names(x))
  x
})

## merge all data by cvd_studyid
phs.aim2.data = Reduce(function(...) merge(...,by='cvd_studyid', all.x=T), datalist)

phs.aim2.data$cvd_condition = cvd$cvd_condition[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$cvd_group = cvd$cvd_group[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$daysto_cvd_event = cvd$daysto_cvd_event[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]

phs.aim2.data$daysto_cvd_dt = phs.aim2.data$index_date + phs.aim2.data$daysto_cvd_event


# format and create new variables
# age groups
phs.aim2.data$agegrp = cut(phs.aim2.data$dxage, c(0, 40, 50, 60, 70, 101), right=F,
                           labels = c('<40 yo','40-49','50-59','60-69','70+ yo'))

# race
phs.aim2.data$race = factor(phs.aim2.data$raceethn1, 
                            labels=c("WHITE","BLACK","ASIAN","HISPANIC","PI","AI-AN"))

# postive nodes
phs.aim2.data$nodal = factor(phs.aim2.data$nodal, labels = c('Positive','Negative'))

# tumor markers
phs.aim2.data[,c('er','pr','her2')] = lapply(phs.aim2.data[,c('er','pr','her2')], function(x)
  factor(x, levels = c(0,1,2,3,8,9),
         labels = c('Not done','Positive','Negative',
                    'Borderline','Ordered, N/A',
                    'Unknown')))

# erpr status
phs.aim2.data$erpr = factor(phs.aim2.data$erpr, labels = c('ER+/PR+','ER+/PR-','ER-/PR+','ER-/PR-',
                                                           'UNKNOWN'))

# more tumor markers and treatment received
phs.aim2.data[,c('tri_neg','chemo_yn','rad_tx_yn','horm_yn')] = 
  lapply(phs.aim2.data[,c('tri_neg','chemo_yn','rad_tx_yn','horm_yn')], function(x)
    factor(x, levels = c(0,1,9),
           labels = c('No','Yes','Other')))

# AJCC stage
phs.aim2.data$ajcc_stage = factor(phs.aim2.data$ajcc_stage,
                                  labels = c('Stage I','Stage II', 'Stage III','Stage IV'))


# BMI category
phs.aim2.data$bmi1 = as.numeric(gsub('[<|>|=]','',phs.aim2.data$bmi))
phs.aim2.data$bmicat = cut(phs.aim2.data$bmi1, c(0,18.5,25,30,35,Inf),right=F,
                           labels = c('Underweight','Normal','Overweight','Obese I','Obese II+'))
phs.aim2.data$bmicat = as.character(phs.aim2.data$bmicat)
phs.aim2.data$bmicat[phs.aim2.data$raceethn1=='ASIAN' & (phs.aim2.data$bmi1 >= 23 & phs.aim2.data$bmi1 < 27.5)] = 'Overweight'
phs.aim2.data$bmicat[phs.aim2.data$raceethn1=='ASIAN' & (phs.aim2.data$bmi1 >= 27.5 & phs.aim2.data$bmi1 < 35)] = 'Obese I'
phs.aim2.data$bmicat[is.na(phs.aim2.data$bmicat)] = 'Unknown'
phs.aim2.data$bmicat = factor(phs.aim2.data$bmicat,
                              levels = c('Underweight','Normal','Overweight','Obese I','Obese II+','Unknown'))

# smoking status
phs.aim2.data$smok = phs.aim2.data$smoke_status_6m
phs.aim2.data$smok[phs.aim2.data$smok %in% c('Q')] = 'Quited'
phs.aim2.data$smok[phs.aim2.data$smok %in% c('Y')] = 'Current smoker'
phs.aim2.data$smok[phs.aim2.data$smok %in% c('N')] = 'Never smoker'
phs.aim2.data$smok[(phs.aim2.data$smok %in% c(''))|is.na(phs.aim2.data$smok)] = 'Unknown'

phs.aim2.data$smok = factor(phs.aim2.data$smok, levels=c('Never smoker','Current smoker','Quited','Unknown'))

# menopausal status
phs.aim2.data$menop = factor(phs.aim2.data$bl_meno_status)
phs.aim2.data$menop[is.na(phs.aim2.data$menop) & phs.aim2.data$age > 51] = '1'
phs.aim2.data$menop[is.na(phs.aim2.data$menop) & phs.aim2.data$age <= 51] = '0'

# baseline diabetes                                                 
phs.aim2.data$diab_bl = ifelse(phs.aim2.data$daysto_diabetes<=0, 1, 0)
phs.aim2.data$diab_bl[is.na(phs.aim2.data$diab_bl)] = 0
phs.aim2.data$diab_bl = factor(phs.aim2.data$diab_bl,levels=c(1,0),labels = c('Yes','No'))

# baseline dyslipidemia
phs.aim2.data$dyslipid_bl = ifelse(phs.aim2.data$daysto_dyslipidemia<=0, 1, 0)
phs.aim2.data$dyslipid_bl[is.na(phs.aim2.data$dyslipid_bl)] = 0
phs.aim2.data$dyslipid_bl = factor(phs.aim2.data$dyslipid_bl,levels=c(1,0),labels = c('Yes','No'))

# baseline hypertension
phs.aim2.data$htn_bl = ifelse(phs.aim2.data$daysto_phase_htn<=0, 1, 0)
phs.aim2.data$htn_bl[is.na(phs.aim2.data$htn_bl)] = 0
phs.aim2.data$htn_bl = factor(phs.aim2.data$htn_bl,levels=c(1,0),labels = c('Yes','No'))

# education
phs.aim2.data$edu_cat = educ_income$BL_EDUCLVL_5CAT[match(phs.aim2.data$cvd_studyid,educ_income$CVD_studyid)]
phs.aim2.data$edu_cat = as.factor(phs.aim2.data$edu_cat)

# income
phs.aim2.data$income_cat = educ_income$BL_INC_CAT2[match(phs.aim2.data$cvd_studyid,educ_income$CVD_studyid)]
phs.aim2.data$income_cat = as.factor(phs.aim2.data$income_cat)



### Define CVD outcomes
## set all NA for cvd outcomes as 0
phs.aim2.data[,tolower(c(cvd_grp,cvd_cond))] = lapply(phs.aim2.data[,tolower(c(cvd_grp,cvd_cond))], function(x) {
  x[is.na(x)] = 'No'
  x
})

phs.aim2.data[,grep('inc$|prev$|rec$',names(phs.aim2.data))] = 
  lapply(phs.aim2.data[,grep('inc$|prev$|rec$',names(phs.aim2.data))], function(x) {
    x[is.na(x)] = 0
    x
  })


phs.aim2.data$heart_failure_cardiomyopathy_grp_prev = ifelse(phs.aim2.data$cardiomyopathy_grp_prev==1|
                                           phs.aim2.data$heart_failure_grp_prev==1, 1, 0)

phs.aim2.data$heart_failure_cardiomyopathy_grp_inc =  ifelse(phs.aim2.data$cardiomyopathy_grp_inc==1|
                                                               phs.aim2.data$heart_failure_grp_inc==1, 1, 0)

phs.aim2.data$heart_failure_cardiomyopathy_grp_incdt = apply(phs.aim2.data[,c("cardiomyopathy_grp_incdt",
                                                          "heart_failure_grp_incdt")],1,min,na.rm=T)


# define censoring time
phs.aim2.data$censor_dt = apply(phs.aim2.data[,c("daysto_death","daysto_enr_end","daysto_end_study")],1,min,na.rm=T)


# calculate time of follow up
phs.aim2.data$ischemic_heart_disease_grp_inc_fu = phs.aim2.data$ischemic_heart_disease_grp_incdt
phs.aim2.data$ischemic_heart_disease_grp_inc_fu[which(phs.aim2.data$ischemic_heart_disease_grp_inc==0)] = 
  phs.aim2.data$censor_dt[which(phs.aim2.data$ischemic_heart_disease_grp_inc==0)] 

phs.aim2.data$stroke_grp_inc_fu = phs.aim2.data$stroke_grp_incdt
phs.aim2.data$stroke_grp_inc_fu[which(phs.aim2.data$stroke_grp_inc==0)] = 
  phs.aim2.data$censor_dt[which(phs.aim2.data$stroke_grp_inc==0)]

phs.aim2.data$cardiomyopathy_grp_inc_fu = phs.aim2.data$cardiomyopathy_grp_incdt
phs.aim2.data$cardiomyopathy_grp_inc_fu[which(phs.aim2.data$cardiomyopathy_grp_inc==0)] = 
  phs.aim2.data$censor_dt[which(phs.aim2.data$cardiomyopathy_grp_inc==0)] 

phs.aim2.data$heart_failure_grp_inc_fu = phs.aim2.data$heart_failure_grp_incdt
phs.aim2.data$heart_failure_grp_inc_fu[which(phs.aim2.data$heart_failure_grp_inc==0)] = 
  phs.aim2.data$censor_dt[which(phs.aim2.data$heart_failure_grp_inc==0)] 

phs.aim2.data$heart_failure_cardiomyopathy_grp_inc_fu = phs.aim2.data$heart_failure_cardiomyopathy_grp_incdt
phs.aim2.data$heart_failure_cardiomyopathy_grp_inc_fu[which(phs.aim2.data$heart_failure_cardiomyopathy_grp_inc==0)] = 
  phs.aim2.data$censor_dt[which(phs.aim2.data$heart_failure_cardiomyopathy_grp_inc==0)] 


```

# Define eofdate and eoftype
## EOF_date (date of end of follow-up): earliest time among date of death, end of study date, CVD events date, or end of enrollment date.
## EOF_type (reason for end of follow-up): 

```{r, message=FALSE, warning=FALSE}
phs.aim2.data$death_date = as.Date(phs.aim2.data$death_date)
phs.aim2.data$end_of_study = as.Date(phs.aim2.data$end_of_study)
phs.aim2.data$daysto_cvd_dt = as.Date(phs.aim2.data$daysto_cvd_dt)
phs.aim2.data$enr_end = as.Date(phs.aim2.data$enr_end)
phs.aim2.data$index_date = as.Date(phs.aim2.data$index_date)

phs.aim2.data$eofdate = pmin(phs.aim2.data$death_date, phs.aim2.data$end_of_study, phs.aim2.data$daysto_cvd_dt, phs.aim2.data$enr_end, na.rm = T)

table(phs.aim2.data$eofdate < phs.aim2.data$index_date)

phs.aim2.data$eofdate[phs.aim2.data$eofdate < phs.aim2.data$index_date] = phs.aim2.data$index_date[phs.aim2.data$eofdate < phs.aim2.data$index_date]

# Create eoftype variable
phs.aim2.data$ischemic_heart_disease_event = cvd$ischemic_heart_disease[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$heart_failure_event = cvd$heart_failure[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$cardiomyopathy_event = cvd$cardiomyopathy[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$stroke_event = cvd$stroke[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$tia_event = cvd$tia[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$valvular_disease_event = cvd$valvular_disease[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$myocarditis_pericarditis_event = cvd$myocarditis_pericarditis[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$carotid_disease_event = cvd$carotid_disease[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$cardiac_arrest_event = cvd$cardiac_arrest[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$arrhythmia_event = cvd$arrhythmia[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]
phs.aim2.data$vte_event = cvd$vte[match(phs.aim2.data$cvd_studyid, cvd$cvd_studyid)]


phs.aim2.data$eoftype = NA
phs.aim2.data$eoftype[!is.na(phs.aim2.data$death_date)] = "Death"
phs.aim2.data$eoftype[phs.aim2.data$eofdate == "2018-12-31"] = "End of Study"
phs.aim2.data$eoftype[phs.aim2.data$ischemic_heart_disease_event == 1] = "Ischemic Heart Disease"
phs.aim2.data$eoftype[phs.aim2.data$heart_failure_event == 1 | phs.aim2.data$cardiomyopathy_event == 1] = "Heart Failure/Cardiomyopathy"
#phs.aim2.data$eoftype[phs.aim2.data$cardiomyopathy_event == 1] = "Cardiomyopathy"
phs.aim2.data$eoftype[phs.aim2.data$stroke_event == 1] = "Stroke"
phs.aim2.data$eoftype[phs.aim2.data$arrhythmia_event == 1] = "Arrhythmia"
phs.aim2.data$eoftype[phs.aim2.data$cardiac_arrest_event == 1] = "Cardiac Arrest"
phs.aim2.data$eoftype[phs.aim2.data$carotid_disease_event == 1] = "Carotid Disease"
phs.aim2.data$eoftype[phs.aim2.data$myocarditis_pericarditis_event == 1] = "Myocarditis Pericarditis"
phs.aim2.data$eoftype[phs.aim2.data$tia_event == 1] = "TIA"
phs.aim2.data$eoftype[phs.aim2.data$valvular_disease_event == 1] = "Valvular Disease"
phs.aim2.data$eoftype[phs.aim2.data$vte_event == 1] = "VTE"
phs.aim2.data$eoftype[is.na(phs.aim2.data$eoftype)] = "Disenrollment"

table(phs.aim2.data$eoftype)


```



